---
title: "Introduction to gerda"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gerda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gerda)
```

## Overview

The `gerda` package provides functions to access and work with GERDA datasets. The German Election Database (GERDA) provides a comprehensive dataset of local, state, and federal election results in Germany. The data is intended to facilitate research on electoral behavior, representation, and political responsiveness at multiple levels of government. All datasets include turnout and vote shares for all major parties. Moreover, GERDA contains geographically harmonized datasets that account for changes in municipal boundaries and mail-in voting districts.

GERDA was compiled by Vincent Heddesheimer, Florian Sichart, Andreas Wiedemann and Hanno Hilbig.

This vignette will introduce you to the main functions of the package and demonstrate how to use them.

## Available Datasets

To see a list of all available GERDA datasets, you can use the `gerda_data_list()` function:

```{r}
gerda_data_list()
```

This function displays a formatted table with the names and descriptions of all available datasets.

## Loading Data

The main function for loading GERDA data is `load_gerda_web()`. This function allows you to load a specific dataset from a web source. Here's an example of how to use it:

```{r, eval=FALSE}
# Load the municipal harmonized dataset
municipal_harm_data <- load_gerda_web("municipal_harm", verbose = TRUE, file_format = "rds")
```

The `load_gerda_web()` function takes the following parameters:

- `file_name`: The name of the dataset to load (as shown in the `gerda_data_list()` output)
- `verbose`: If set to `TRUE`, it prints messages about the loading process (default is `FALSE`)
- `file_format`: Specifies the format of the file to load, either "rds" or "csv" (default is "rds")

## Example Workflow

Here's an example of a typical workflow using the `gerda` package:

1. List available datasets:

```{r}
gerda_data_list()
```

2. Load a dataset (in this case, the federal elections at the county level, harmonized):

```{r, eval=FALSE}
federal_cty_harm <- load_gerda_web("federal_cty_harm", verbose = TRUE)
```


## Conclusion

The `gerda` package provides easy access to a wide range of German election and related data. By using the `gerda_data_list()` function to explore available datasets and `load_gerda_web()` to load them, you can quickly incorporate this data into your research or analysis projects.

For more information or to provide feedback, please contact <hhilbig@ucdavis.edu> or visit the GitHub repository at https://github.com/hhilbig/gerda.
