---
title: "Introduction to gerda"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gerda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gerda)
```

## Overview

The `gerda` package provides functions to access and work with GERDA datasets. The German Election Database (GERDA) provides data on federal elections, state (*Landtag*) elections, local (*Kommunal*) elections, mayoral (*Bürgermeister*) elections, Landrat elections, European Parliament elections, and county (*Kreistag*) elections. Federal data include municipality-, county-, and constituency-level files; state data include municipality- and constituency-level files. GERDA also supplies geographically harmonized datasets that account for changes in municipal boundaries and mail-in voting districts.

In addition to election results, the package provides county-level socioeconomic covariates from INKAR, municipality-level data from the German Census 2022, and a party crosswalk that maps GERDA party names to standardized ParlGov attributes.

GERDA was compiled by Vincent Heddesheimer, Florian Sichart, Andreas Wiedemann and Hanno Hilbig. For additional information, see also the GERDA website (www.german-elections.com) and the accompanying publication: [doi.org/10.1038/s41597-025-04811-5](https://doi.org/10.1038/s41597-025-04811-5)

This vignette will introduce you to the main functions of the package and demonstrate how to use them. 

For a stricter workflow intended for unattended coding agents, including
snapshot provenance, key audits, guarded joins, and machine-readable join
diagnostics, see `vignette("agent-workflow")`.

## Available Datasets

To see a list of all available GERDA electoral result datasets, you can use the `gerda_data_list()` function:

```{r}
gerda_data_list()
```

This function displays a formatted table with the names and descriptions of all available datasets. Use `gerda_data_list(print_table = FALSE)` to obtain the structured catalog, then pass a value from its `data_name` column to `load_gerda_web()`.

## Loading Data

The main function for loading GERDA data is `load_gerda_web()`. This function allows you to load a specific dataset from a web source. Here's an example of how to use it:

```{r, eval=FALSE}
# Load the municipal harmonized dataset
municipal_harm_data <- load_gerda_web(
  "municipal_harm",
  verbose = TRUE,
  file_format = "rds",
  on_error = "stop",
  cache = TRUE
)
```

The `load_gerda_web()` function takes the following parameters:

- `file_name`: A character string with the name of the dataset to load, e.g. `"federal_cty_harm"` (as shown in the `gerda_data_list()` output). The function supports fuzzy matching, so close misspellings will produce a helpful suggestion.
- `verbose`: If set to `TRUE`, it prints messages about the loading process (default is `FALSE`)
- `file_format`: Specifies the format of the file to load, either `"rds"` or `"csv"` (default is `"rds"`). Both formats return the same tibble, so this choice only affects download size and speed.
- `on_error`: Use `"stop"` in scripts and pipelines so a failed load raises an immediate error instead of returning `NULL` (the backwards-compatible default is `"warn"`).
- `timeout` and `max_retries`: Control the download timeout and number of retry attempts.
- `cache`: If `TRUE`, retain a local copy for subsequent calls. The default is `FALSE`.
- `refresh`: If `TRUE`, ignore a cached copy and download the dataset again.

## Choosing the Right Dataset

The dataset variant is a substantive choice:

- `*_raw` files retain a source-near representation and are mainly useful for auditing or custom cleaning.
- `*_unharm` files use the geography in force at each election. They are suitable for cross-sectional analyses, but geographic identifiers can refer to changing territories over time.
- `*_harm` files map observations to a common geography for comparisons over time. A suffix such as `_21`, `_23`, or `_25` states the target boundary vintage. Select the vintage that matches the analysis rather than automatically choosing the newest one.
- Wahlkreis files ending in `_long` contain party vote counts; their wide counterparts contain party vote shares. State and federal Wahlkreis boundaries can change between elections, and the unharmonized files do not by themselves create fixed-boundary panels.

The structured catalog supports programmatic filtering:

```{r}
catalog <- gerda_data_list(print_table = FALSE)
subset(catalog,
       election_type == "federal" &
       geographic_level == "municipality" &
       boundary == "2025")
```

## Example Workflow

Here's an example of a typical workflow using the `gerda` package:

1. List available datasets:

```{r}
gerda_data_list()
```

2. Load a dataset (in this case, the federal elections at the county level, harmonized):

```{r, eval=FALSE}
federal_cty_harm <- load_gerda_web(
  "federal_cty_harm",
  verbose = TRUE,
  on_error = "stop",
  cache = TRUE
)
```

## Joining GERDA Datasets

If you are using `add_gerda_covariates()` or `add_gerda_census()`, you can skip this section: the helpers detect the level of your data and use the correct join keys automatically. If you are writing a manual `left_join()` or merging against other sources, the table below shows which identifier and time columns each family carries.

| Dataset family | Geographic identifier and additional row keys | Time column |
|---|---|---|
| `municipal_*`, `state_*`, `federal_muni_*`, `european_muni_*` | `ags` (8-digit municipality) | `election_year` (+ `election_date` where available) |
| `mayoral_unharm`, `mayoral_harm` | `ags`; `round` distinguishes first and runoff rounds | `election_year`, `election_date` |
| `mayoral_candidates` | `ags` + candidate; first- and second-round results are stored in separate columns | `election_year`, `election_date`, `election_date_sw` |
| `mayor_panel`, `mayor_panel_harm` | `ags` + `person_id` | `election_date` |
| `mayor_panel_annual`, `mayor_panel_annual_harm` | `ags` + `person_id` | `year` |
| `landrat_unharm` | `ags` (8-digit county AGS ending in `000`); `round` distinguishes election rounds | `election_year`, `election_date` |
| `landrat_candidates` | `ags` (8-digit county AGS ending in `000`) + candidate | `election_year`, `election_date`, `election_date_sw` |
| `federal_cty_harm` | `county_code` (5-digit county) | `election_year` |
| `federal_cty_unharm` | `county_code` (renamed from upstream `ags` on load) | `election_year` (renamed from upstream `year` on load) |
| `county_elec_*_cty` | `county_code` (5-digit county) | `election_year` |
| `county_elec_unharm`, `county_elec_*_muni` | `ags` (8-digit municipality) | `election_year` |
| `county_council_seats` | `county` (5-digit county) | `year` (yearly panel; seat vectors carried forward between elections) |
| `ltw_wkr_unharm` | `state` + `wkr_nr` + `stimme` | `election_year`, `election_date` |
| `ltw_wkr_unharm_long` | `state` + `wkr_nr` + `stimme` + `party` | `election_year`, `election_date` |
| `federal_wkr_unharm` | `wkr_nr` + `stimme` | `election_year`, `election_date` |
| `federal_wkr_unharm_long` | `wkr_nr` + `stimme` + `party` | `election_year`, `election_date` |
| `federal_wkr_2021_on_2025` | `wkr_nr` (2025 boundaries) + `stimme` | Fixed 2021 result; no time column |
| `wkr_2021_to_2025_crosswalk` | `wkr_nr` (2025) + `prior_2021_wkr_nr` | Boundary vintages are encoded in column names |
| `gerda_covariates()` (INKAR, county-level) | `county_code` (5-digit county) | `year` (not `election_year`) |
| `gerda_census()` (Zensus 2022, municipality-level) | `ags` (8-digit municipality) | *Time-invariant (2022)* |
| `ags_crosswalks`, `ags_1990_to_2023_crosswalk`, `ags_1990_to_2025_crosswalk`, `crosswalk_ags_*` | Pair of AGS codes at source and target vintages | Vintage is encoded in column names |
| `cty_crosswalks` | Pair of 5-digit county codes at source and target vintages | Vintage is encoded in column names |

Three things to watch for when joining manually:

- `gerda_covariates()` uses `year`, not `election_year`. If you merge it directly into federal election data, you need a rename: `by = c("county_code" = "county_code", "election_year" = "year")`.
- Keep AGS and Wahlkreis identifiers as character strings so leading zeros are preserved.
- On municipal-level files, the `county` column is a county name or partial code, not the 5-digit AGS-based county code. Use `substr(ags, 1, 5)` to extract the county key when you want to join against county-level data.

## County-Level Covariates

The `gerda` package includes county-level socioeconomic and demographic covariates from INKAR (Indikatoren und Karten zur Raum- und Stadtentwicklung). These covariates can be easily merged with GERDA election data to enrich your analyses. INKAR data is available from 1995 to 2022, so covariates can be matched to federal elections from 1998 onwards (earlier elections fall outside the INKAR coverage window).

### Quick Start

The easiest way to add covariates to your election data is using the `add_gerda_covariates()` function:

```{r, eval=FALSE}
library(dplyr)

# Load election data and add covariates
merged <- load_gerda_web(
  "federal_cty_harm",
  on_error = "stop",
  cache = TRUE
) %>%
  add_gerda_covariates(unmatched = "error")

join_report <- gerda_join_diagnostics(merged)
```

Under the hood, `add_gerda_covariates()` merges on county code and election year. It automatically:

- Detects whether the input is county-level or municipal-level data (and extracts county codes from municipal AGS codes if needed)
- Performs a left join, so all election rows are kept and covariates are added where available
- Validates that the required columns (`county_code` or `ags`, and `election_year`) are present
- Rejects malformed or numeric identifiers, duplicate reference keys, output-column conflicts, and row expansion
- Reports unexpected unmatched keys separately from election years outside the 1995-2022 INKAR coverage window

The default `unmatched = "warn"` retains unexpected unmatched rows with missing
joined values and reports exact row and unit counts. Use `"error"` in automated
pipelines or `"ignore"` only when unmatched keys have already been audited.
Retrieve the machine-readable report with `gerda_join_diagnostics()`.

### Available Covariates

The covariates dataset includes 30 variables across 10 categories (for the full list of variable names, units, and descriptions, see `gerda_covariates_codebook()`):

- **Demographics**: Age structure, foreign population, gender composition
- **Economy**: GDP per capita, sectoral composition, enterprise structure
- **Labor Market**: Unemployment rates (overall, youth, long-term)
- **Education**: School completion rates, students, apprentices
- **Income**: Purchasing power, low-income households
- **Healthcare**: Physician density, hospital beds, GP density
- **Childcare**: Coverage rates for under-3 and 3-6 age groups
- **Housing**: Building permits, rent levels, living space
- **Transport**: Cars per capita
- **Public Finances**: Municipal debt, tax revenue

### Viewing the Codebook

To see detailed information about each covariate, including units and missing data patterns:

```{r, eval=FALSE}
# Get the codebook
codebook <- gerda_covariates_codebook()
print(codebook)

# Find variables with good coverage
library(dplyr)
codebook %>%
  filter(missing_pct < 10) %>%
  select(variable, label, category)
```

### Advanced Usage

For more control, you can access the raw covariates data:

```{r, eval=FALSE}
# Get raw covariate data
covs <- gerda_covariates()

# Inspect before merging
summary(covs$unemployment_rate)

# Custom merge
elections <- load_gerda_web("federal_cty_harm")
merged <- elections %>%
  left_join(covs, by = c("county_code" = "county_code", "election_year" = "year"))
```

### Data Coverage

- **Counties**: 400 German counties (Kreise)
- **Time period**: 1995-2022 (annual data)
- **Election coverage**: Elections from 1998 onwards have covariate data

Coverage varies by variable: core indicators (demographics, economy, labor market) are available for all 7 federal election years (1998-2021). Newer INKAR indicators (e.g., childcare, some healthcare variables) are available for 2-3 recent elections only. Consult the codebook's `missing_pct` column to check per-variable availability before analysis.

## Census 2022 Data

The `gerda` package includes municipality-level data from the German Census 2022 (Zensus 2022). This cross-sectional snapshot covers approximately 10,800 municipalities and can be merged with any GERDA election dataset.

The main advantage of this covariate data is that it is observed at the municipal level (unlike the county-level INKAR data). This allows for more fine-grained analyses of local election outcomes. However, the census is a single time point (2022), so it does not vary across election years. This means that the resulting merged dataset will have time-invariant covariates, i.e. each municipality receives the same census values for all election years. Users should not conduct analyses that rely on over-time variation in these covariates.

### Quick Start

```{r, eval=FALSE}
library(gerda)

# Add census data to municipal-level elections
muni_merged <- load_gerda_web(
  "federal_muni_harm_21",
  on_error = "stop",
  cache = TRUE
) |>
  add_gerda_census(unmatched = "error")

# Also works with county-level data (aggregated from municipalities)
county_merged <- load_gerda_web(
  "federal_cty_harm",
  on_error = "stop",
  cache = TRUE
) |>
  add_gerda_census(unmatched = "error")

gerda_join_diagnostics(muni_merged)
```

### Available Indicators

The census data includes 14 indicators across four categories:

- **Demographics**: Population, age structure (under 18, 18-29, 30-49, 50-59, 60+)
- **Migration**: Migration background share, foreign nationals share
- **Households**: Average household size
- **Housing**: Total dwellings, vacancy rate, ownership rate, average rent per m², single-family home share

Since the census is a 2022 snapshot, the same values are attached to all election years (see also the note above). 

The Destatis source publishes a combined 60-74 age bin. GERDA therefore exposes `share_50to59_census22` and `share_60plus_census22`; the source cannot support separate 50-64 and 65+ measures.

### Viewing the Codebook

```{r, eval=FALSE}
# Get the census codebook
census_cb <- gerda_census_codebook()
print(census_cb)
```

### Data Coverage

Most census variables have >95% municipality coverage. `avg_household_size_census22` has approximately 12.5% missing values because Destatis suppresses data for small municipalities under its disclosure rules.

## Party Crosswalk Function

The `party_crosswalk()` function provides a mapping between GERDA party names and standardized party information from the ParlGov database. This is particularly useful for linking GERDA data with other political science datasets or for obtaining standardized party characteristics.

### Usage

The function takes two main parameters:

- `party_gerda`: A character vector of GERDA party names
- `destination`: The name of the column from the ParlGov view_party table to map to

### Available Mapping Options

You can map GERDA party names to various standardized party characteristics, including:

- `left_right`: Left-right position scores
- `party_name_english`: English party names
- `party_name_short`: Short party names
- `country_name`: Country names
- And many other ParlGov variables

### Example

```{r, eval=FALSE}
# Map GERDA party names to left-right positions
parties <- c("cdu", "spd", "linke_pds", "fdp")
left_right_scores <- party_crosswalk(parties, "left_right")
print(left_right_scores)

# Map to English party names
english_names <- party_crosswalk(parties, "party_name_english")
print(english_names)
```

This function is especially useful when you want to:

- Analyze parties along ideological dimensions
- Merge GERDA data with other comparative datasets
- Standardize party names across different data sources
- Access additional party metadata from ParlGov

## Conclusion

The `gerda` package provides easy access to a wide range of German election and related data. By using the `gerda_data_list()` function to explore available datasets and `load_gerda_web()` to load them, you can quickly incorporate this data into your research or analysis projects.

For more information or to provide feedback, please contact <hhilbig@ucdavis.edu> or visit the GitHub repository at https://github.com/hhilbig/gerda.
