---
title: "Introduction to gerda"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gerda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gerda)
```

## Overview

The `gerda` package provides functions to access and work with GERDA datasets. This vignette will introduce you to the main functions of the package and demonstrate how to use them.

## Available Datasets

To see a list of all available GERDA datasets, you can use the `gerda_data_list()` function:

```{r}
gerda_data_list()
```

This function displays a formatted table with the names and descriptions of all available datasets.

## Loading Data

The main function for loading GERDA data is `load_gerda_web()`. This function allows you to load a specific dataset from a web source. Here's an example of how to use it:

```{r, eval=FALSE}
# Load the municipal harmonized dataset
municipal_harm_data <- load_gerda_web("municipal_harm", verbose = TRUE, file_format = "rds")
```

The `load_gerda_web()` function takes the following parameters:

- `file_name`: The name of the dataset to load (as shown in the `gerda_data_list()` output)
- `verbose`: If set to `TRUE`, it prints messages about the loading process (default is `FALSE`)
- `file_format`: Specifies the format of the file to load, either "rds" or "csv" (default is "rds")

## Example Workflow

Here's an example of a typical workflow using the `gerda` package:

1. List available datasets:

```{r}
gerda_data_list()
```

2. Load a dataset (in this case, the federal elections at the county level, harmonized):

```{r, eval=FALSE}
federal_cty_harm <- load_gerda_web("federal_cty_harm", verbose = TRUE)
```


## Conclusion

The `gerda` package provides easy access to a wide range of German election and related data. By using the `gerda_data_list()` function to explore available datasets and `load_gerda_web()` to load them, you can quickly incorporate this data into your research or analysis projects.

For more information or to provide feedback, please contact <hhilbig@ucdavis.edu> or visit the GitHub repository at https://github.com/hhilbig/gerda.
